Unifying Lower Bounds on Prediction Dimension of Consistent Convex Surrogates

### 2021-05-24 cvx-flats-neurips21/cvx-flats.tex
Pass over 3.1 and adding Corollary text after Lemma 1

@@

\paragraph{Q1: Classification with an abstain option.}
The abstain target loss is a well-studied variation of 0-1 loss that allows for an ``abstain'' report that gives a lesser punishment $1/2$ for abstaining, $r = \bot$ \raft{cites here}.
Formally, the target loss is $\ell^{1/2}(r,y) := \Ind{r \not \in \{y, \bot\}} + (1/2)\Ind{r = \bot}$.
Since we are given a discrete target loss, this problem fits nicely into Quadrant 1.

To apply Lemma~\ref{lem:convex-flats-inf-dim}, we first consider the abstain property $\gamma$ elicited by $\ell^{1/2}$, where one predicts the most likely outcome $y$ if $Pr[Y=y] \geq 1/2$ and otherwise ``abstains'' by predicting $\bot$.
% Upon closer inspection, the level set boundaries are found at the set of distribution where is a $y$ such that $p_y = 1/2$.
For the depicted distribution $p \in \relint{\gamma_\bot}$, we can derive a lower bound on convex consistency dimension as we cannot fit a $1$-flat (line) fully contained in $\gamma_\bot$ that passes through $p$.
By Corollary~\ref{cor:Pcodim-flat-elic-relint-prop}, we can conclude $\conscvx(\gamma^{1/2}) \geq 2$ when $|\Y|=3$, which matches the upper bound from the convex surrogate of
\citet{ramaswamy2016convex}.\raft{Jessie: see if you can match the style of the revised abstain example for the others.  In particular, I think we can leave the definition of $p$ to the figure, and avoid referencing $\bullet$ or anything -- the figure can make clear that $p=\bullet$.  Probably it's only Q2 that needs to change.}
%present a convex surrogate for the abstain loss yielding a logarithmic upper bound on $\conscvx(\ell^{1/2})$, so this bound is tight.% which are an exponential improvement over previous results, e.g.,~\citep{crammer2001algorithmic,rifkin2004defense}.

\paragraph{Q2: High confidence classification.}
Consider the following scenario where someone is deciding how to dress for the weather based on a meteorologist's forecast: a forecaster is choosing between three outcomes $\Y = \{$rainy, sunny, snowy$\}$, and we suppose we want to have some bias towards health and safety, so the meteorologist should only predict sunny weather if $Pr[$sunny $|$ weather data$] \geq 3/4$.
Otherwise, they should predict whatever is more likely  given the weather data: rain or snow.

@@

This property is shown in Table~\ref{tab:quadrants-simplices} (Q2), and we can observe the point $p = \bullet = (0.3,0.3,0.4)$ that, like abstain, we cannot fit a $1$-flat entirely contained in the level set $\gamma_{\text{snowy}}$ through $p$.
Therefore, we can apply Corollary~\ref{cor:Pcodim-flat-single-val-prop} to conclude that $\conscvx(\gamma) = 2$.

\paragraph{Q3: Least-squares regression}
\jessie{Define property, discuss briefly why in the quadrant.  Move to how results apply and what they say}
While least squares regression is commonly used in machine learning and statistics for continuous estimation, it might not be immediately \emph{why} we do so, prompting us to study this smooth convex loss in Quadrant 3.
Evaluating the derivative of squared loss, we can observe that it elicits the expected value $\Gamma(p) = \E_p[Y]$.
With $\Y = \{-1,0,1\}$, we can visualize the level sets of the expected value as in Table~\ref{tab:quadrants-simplices} (Q3).
Since squared loss is a $1$-dimensional loss, we can verify that for all $p \in \simplex$, we can fit a $1$-flat (precisely equal to the level set) through $p$, as demonstrated in the Table.

\paragraph{Q4: Variance}
In understanding uncertainty, one might reasonably want to estimate the variance $\Gamma^\Var(p) = \E_p[Y^2] - \E_p[Y]^2$.
This property is continuous in the simplex $\simplex$, and therefore belongs in Quadrant 4.
The well-known fact that the variance is not elicitable does not yield a lower bound on elicitation complexity of 2, as it does not rule out the variance being a link of a real-valued convex-elicitable property; cf.~\citet[Remark 1]{frongillo2020elicitation}.
However, in Appendix~\ref{app:omitted-examples}, we show $\eliccvx(\Gamma^\Var)=2$, meaning the lowest dimension of a convex loss to estimate conditional variance is 2. %~\citep{osband1985providing,lambert2018elicitation}.
This lower bound will follow from Theorem~\ref{thm:bayes-risk-lower-bound} using the fact that variance is the Bayes risk of squared loss $L(r,y) = (r-y)^2$, which elicits the mean $\Gamma(p) = \E_p Y$.
Interestingly, while perhaps intuitively obvious, even this simple result is novel.

@@

Our Lemma~\ref{lem:convex-flats-inf-dim} subsumes the  bounds given by~\citet{ramaswamy2016convex} by showing that, if there is a $d$-flat fully contained in the level set (so we can apply Lemma~\ref{lem:convex-flats-inf-dim}) then the co-dimension of the subspace of feasible directions at the same $p \in \C := \gamma_r$ has co-dimension at most $d$.

@@

\begin{figure}[htb]
	\centering
	\begin{tabular}{cc}
%				\includegraphics[width=.35\textwidth]{tikz/fsd-bound.pdf} & 
%				\includegraphics[width=.35\textwidth]{tikz/flats-bound.pdf} \\
	\end{tabular}
	\caption{(Left) Feasible subspace dimension $\Sc_{\gamma_\bot}(\bullet) = 2$ and $\Sc_{\gamma_\bot}({\color{blue}\star}) = 1$, giving the bound $\conscvx(\ell^{abs}) \geq 3- 1-1 = 1$.
		(Right) No $1$-flat through $\bullet$ (a line since $\bullet \in \relint{\simplex})$  stays fully contained in $\gamma_\bot$, so $\conscvx(\ell^{abs}) \geq 2$.\jessie{FIX FIGURES}}
	\label{fig:fsd-flats-abstain}
\end{figure}

>>>>

Lemma~\ref{lem:convex-flats-inf-dim} now allows us to derive bounds on convex consistency dimension by considering distributions and property values that are either single-valued (Corollary~\ref{cor:Pcodim-flat-single-val-prop}) or on the relative interior of the simplex with finite $\Y$ (Corollary~\ref{cor:Pcodim-flat-elic-relint-prop}).
Proofs are deferred to Appendix~\ref{app:implications-of-bounds}.

@@

\paragraph{Q1: Classification with an abstain option.}
The abstain target loss is a well-studied variation of 0-1 loss that allows for an ``abstain'' report that gives a lesser punishment $1/2$ for abstaining, $r = \bot$~\citep{ramaswamy2018consistent,ramaswamy2016convex,madras2018predict,chow1957optimum,cortes2016learning}.
Formally, the target loss is $\ell^{1/2}(r,y) := \Ind{r \not \in \{y, \bot\}} + (1/2)\Ind{r = \bot}$.
Since we are given a discrete target loss, this problem fits nicely into Quadrant 1.

To apply Lemma~\ref{lem:convex-flats-inf-dim}, we first consider the abstain property $\gamma$ elicited by $\ell^{1/2}$, where one predicts the most likely outcome $y$ if $Pr[Y=y] \geq 1/2$ and otherwise ``abstains'' by predicting $\bot$.
% Upon closer inspection, the level set boundaries are found at the set of distribution where is a $y$ such that $p_y = 1/2$.
For the depicted distribution $p \in \relint{\gamma_\bot}$, we can derive a lower bound on convex consistency dimension as we cannot fit a $1$-flat (line) fully contained in $\gamma_\bot$ that passes through $p$.
By Corollary~\ref{cor:Pcodim-flat-elic-relint-prop}, we can conclude $\conscvx(\gamma^{1/2}) \geq 2$ when $|\Y|=3$, which matches the upper bound from the convex surrogate of
\citet{ramaswamy2016convex}.\raft{Jessie: see if you can match the style of the revised abstain example for the others.  In particular, I think we can leave the definition of $p$ to the figure, and avoid referencing $\bullet$ or anything -- the figure can make clear that $p=\bullet$.  Probably it's only Q2 that needs to change. \jessie{Took a pass! Feel free to burn}}
%present a convex surrogate for the abstain loss yielding a logarithmic upper bound on $\conscvx(\ell^{1/2})$, so this bound is tight.% which are an exponential improvement over previous results, e.g.,~\citep{crammer2001algorithmic,rifkin2004defense}.

\paragraph{Q2: High confidence classification.}
Consider the following scenario where someone is deciding how to dress for the weather based on a meteorologist's forecast: a forecaster is choosing between three outcomes $\Y = \{$rainy, sunny, snowy$\}$, and we suppose we want to have some bias towards health and safety, so the meteorologist should only predict sunny weather if $Pr[$sunny $|$ weather data$] \geq 3/4$.
Otherwise, they should predict whatever is more likely  given the weather data: rain or snow.
\citet{osband1985information-eliciting} notes that a property is elicitable if and only if its level sets form a power diagram projected onto the simplex, so there is no discrete target loss eliciting the described property, hence it belonging in Q2.

@@

This property is shown in Table~\ref{tab:quadrants-simplices} (Q2), and we can observe the point that, like abstain, we cannot fit a $1$-flat (line) entirely contained in the level set $\gamma_{\text{snowy}}$ through $p$.
Therefore, we can apply Lemma~\ref{lem:convex-flats-inf-dim} to conclude that $\conscvx(\gamma) = 2$.

\paragraph{Q3: Least-squares regression}
%\jessiet{Define property, discuss briefly why in the quadrant.  Move to how results apply and what they say}
While least squares regression is commonly used in machine learning and statistics for continuous estimation, it might not be immediately \emph{why} we do so, prompting us to study this smooth convex loss in Quadrant 3.
Evaluating the derivative of squared loss, we observe that it elicits the expected value $\Gamma(p) = \E_p[Y]$.
With $\Y = \{-1,0,1\}$, we can visualize the level sets of the expected value as in Table~\ref{tab:quadrants-simplices} (Q3).
Since squared loss is a $1$-dimensional loss, we can verify that for all $p \in \simplex$, we can fit a $1$-flat (precisely equal to the level set) through $p$, as demonstrated in Table~\ref{tab:quadrants-simplices}.

\paragraph{Q4: Variance}
One might reasonably want to estimate the variance $\Gamma^\Var(p) = \E_p[Y^2] - \E_p[Y]^2$, though a loss function to elicit the variance is not immediately obvious.
This property is continuous in $\simplex$, and therefore belongs in Quadrant 4.
The well-known fact that the variance is not elicitable does not yield a lower bound on elicitation complexity of 2, as it does not rule out the variance being a link of a real-valued convex-elicitable property; cf.~\citet[Remark 1]{frongillo2020elicitation}.
However, in Appendix~\ref{app:omitted-examples}, we show $\eliccvx(\Gamma^\Var)=2$, meaning the lowest dimension of a convex loss to estimate conditional variance is 2. %~\citep{osband1985providing,lambert2018elicitation}.
This lower bound will follow from Theorem~\ref{thm:bayes-risk-lower-bound} in Section~\ref{sec:contin-consis} using the fact that variance is the Bayes risk of squared loss $L(r,y) = (r-y)^2$, which elicits the mean $\Gamma(p) = \E_p Y$.
Interestingly, while perhaps intuitively obvious, even this simple result is novel.

@@

Lemma~\ref{lem:convex-flats-inf-dim} subsumes the  bounds given by~\citet{ramaswamy2016convex} by showing that, if there is a $d$-flat fully contained in the level set (so we can apply Lemma~\ref{lem:convex-flats-inf-dim}) then the co-dimension of the subspace of feasible directions at the same $p \in \C := \gamma_r$ has co-dimension at most $d$.

@@

\begin{figure}[htb]
	\centering
	\begin{tabular}{cc}
%abstain alpha = 1/2
\begin{tikzpicture}
\draw[opacity = 0.2] (2,0) -- (-2,0) -- (0,3.46) -- cycle;
%label outcomes
\node at (-9/4, 0) {$p_1$};
\node at (2/5, 3.25) {$p_2$};
\node at (9/4, 0) {$p_3$};
%level sets
\draw[fill = blue, fill opacity = 0.1, opacity = 0.2] (-1, 1.73) -- (1, 1.73) -- (0, 3.46) -- cycle; \node at (0, 1.73*1.35) {\footnotesize$2$};
\draw[fill = green, fill opacity = 0.1, opacity = 0.2] (2,0) -- (1, 1.73) -- (0,0) -- cycle;
\node at (1, 1.73*0.4) {\footnotesize$3$};
\draw[fill = yellow, fill opacity = 0.1, opacity = 0.2] (-2,0) -- (-1, 1.73) -- (0,0) -- cycle;
\node at (-1, 1.73*0.4) {\footnotesize$1$};
\draw[fill = gray, fill opacity = 0.1, opacity = 0.2] (1,1.73) -- (-1, 1.73) -- (0,0) -- cycle;
\node at (0, 1.73*0.35) {\footnotesize$\bot$};
%uniform distribution
\node at (0, 1.73*0.65) {\textbullet};
\draw[fill=gray, fill opacity = 0.25, dashed] (0,1.73 * 0.65) circle (0.2);

%p = (1/4, 1/4, 1/2)
\draw[thick, dashed, blue] (0, 0) -- (1,1.73);
\node at (0.5, 0.865) {{\color{blue}{\footnotesize$\bigstar$}}};

\end{tikzpicture}
&
%abstain alpha = 1/2
\begin{tikzpicture}
\draw[opacity = 0.2] (2,0) -- (-2,0) -- (0,3.46) -- cycle;
%label outcomes
\node at (-9/4, 0) {$p_1$};
\node at (2/5, 3.25) {$p_2$};
\node at (9/4, 0) {$p_3$};
%level sets
\draw[fill = blue, fill opacity = 0.1, opacity = 0.2] (-1, 1.73) -- (1, 1.73) -- (0, 3.46) -- cycle; \node at (0, 1.73*1.35) {\footnotesize$2$};
\draw[fill = green, fill opacity = 0.1, opacity = 0.2] (2,0) -- (1, 1.73) -- (0,0) -- cycle;
\node at (1, 1.73*0.4) {\footnotesize$3$};
\draw[fill = yellow, fill opacity = 0.1, opacity = 0.2] (-2,0) -- (-1, 1.73) -- (0,0) -- cycle;
\node at (-1, 1.73*0.4) {\footnotesize$1$};
\draw[fill = gray, fill opacity = 0.1, opacity = 0.2] (1,1.73) -- (-1, 1.73) -- (0,0) -- cycle;
\node at (0, 1.73*0.35) {\footnotesize$\bot$};
%uniform distribution
\node at (0, 1.15333) {\textbullet};
\draw[loosely dotted,thick, opacity=0.6] (0, 0) -- (0,3.46);
\draw[loosely dotted,red, thick] (-1.1533, 0) -- (.844945, 2);
%\draw[loosely dotted, blue, thick] (-1.8, 1.15333) -- (1.8, 1.15333);

%p = (1/4, 1/4, 1/2)

\draw[thick,blue,loosely dotted] (0, 0) -- (1,1.73);
\node at (0.5, 0.865) {{\color{blue}{\footnotesize$\bigstar$}}};

\end{tikzpicture}

	\end{tabular}
	\caption{(Left) Feasible subspace dimension $\Sc_{\gamma_\bot}(\bullet) = 2$ and $\Sc_{\gamma_\bot}({\color{blue}\star}) = 1$, giving the bound $\conscvx(\ell^{abs}) \geq 3- 1-1 = 1$.
		(Right) No $1$-flat through $\bullet$ (a line since $\bullet \in \relint{\simplex})$  stays fully contained in $\gamma_\bot$, so $\conscvx(\ell^{abs}) \geq 2$.}
	\label{fig:fsd-flats-abstain}
\end{figure}
